---
title: "Ordinal Logistic Regression"
author: "Gemma Galvez"
date: "`r Sys.Date()`"
toc: true
number-sections: true
format: html
embed-resources: false
echo: true
warning: false
message: false
---

```{r}
#| label: tidyverse
#| echo: true

# Install required packages
library(conflicted)
conflict_prefer("filter", "dplyr", quiet = TRUE)
conflict_prefer("lag", "dplyr", quiet = TRUE)

suppressPackageStartupMessages(library(tidyverse))

# suppress "`summarise()` has grouped output by " messages
options(dplyr.summarise.inform = FALSE)
```

```{r}
#| label: packages
#| echo: true

library(table1)
library(multgee)
library(skimr)
library(pander)
library(gtsummary)
library(car)
library(mltools)
library(MASS)
library(pomcheckr)

conflict_prefer("filter", "dplyr", quiet = TRUE)
conflict_prefer("select", "dplyr", quiet = TRUE)
library(tidyverse)

```

## Introduction

So you’ve found yourself thinking about regression again, haven’t you? I can 
relate. At this point, you’re probably familiar with the fact that logistic 
regression means the logit (log odds) of a binary response is linearly related 
to the independent variable(s). Well, Proportional Odds Logistic Regression is 
an extension of Logistic Regression and that’s likely how you’ve found your way 
here – that, or you really, really enjoy the dark corners of the internet! 
Welcome, make yourself at home.

## Mathematical Definition of Method

Now, [let’s get down to business](https://www.youtube.com/watch?v=TVcLIfSC4OE). Proportional Odds Logistic Regression (POLR), which 
is also known as Ordinal Logistic Regression, is a method of statistical analysis 
which may be used to model the relationship between an ordinal response, and one 
or more independent (or explanatory) variables. These explanatory variables may 
be categorical or continuous. 

Let’s establish some context: Have you ever been filling out a survey and it’s 
asked you how much you agree with something, how often you do something, or how 
important something is to you? While this is not an extensive list, these 
Likert-type responses should give you a general idea of common ordinal responses 
you can expect to encounter when using POLR.   

Some more specific examples where Ordinal Logistic Regression can be applied are: 

* **Level of agreement:** In a survey the responses to the outcome variable is 
categorized in multiple levels such as, Strongly Disagree, Disagree, Agree, 
Strongly Agree.

* **Satisfaction level:** Measuring satisfaction level of a service on a scale 
like, “very dissatisfied,” “dissatisfied,” “neutral,” “satisfied,” and “very 
satisfied.”

* **Pain Intensity:** Patients participating in medical research may be asked to 
rate the intensity of their pain on a scale ranging from “no pain” to “mild pain,” 
“moderate pain,” and “severe pain.”

## Mathematical Formulation of an Ordinal Model

Well, you’ve probably reached the point where you’re about to check out because 
we see formulas, lots of them. This is understandable, but be not afraid! 
[UCLA’s Advanced Research Computing Center](https://stats.oarc.ucla.edu/r/dae/ordinal-logistic-regression/) is here to help in ways I cannot. 

To start… let’s establish some notation and review the concepts involved in 
ordinal logistic regression. Let 𝑌 be an ordinal outcome with 𝐽 categories. 
Then 𝑃(𝑌≤𝑗) is the cumulative probability of 𝑌 less than or equal to a 
specific category 𝑗= 1,⋯,𝐽−1. Note that 𝑃(𝑌≤𝐽)=1. The odds of being less 
than or equal a particular category can be defined as

$$ \frac{P(Y \le j)}{P(Y>j)} $$

For 𝑗=1,⋯,𝐽−1 since 𝑃(𝑌>𝐽)=0 and dividing by zero is undefined. 
Alternatively, you can write 𝑃(𝑌>𝑗)=1–𝑃(𝑌≤𝑗). The log odds is also known 
as the logit, so that

$$ log \frac{P(Y \le j)}{P(Y>j)} = logit (P(Y \le j)) $$

The proportional odds logistic regression model can be defined as

$$logit (P(Y \le j)) = \beta_{j0} – \eta_{1}x_1 – \cdots – \eta_{p} x_p $$

I think that’s enough formulas for now. Let’s keep the party going by actually 
putting the pedal to the metal, or should we say, *aPpLyiNg wHaT wE LeArNeD.* 

## Data Source and Description

The data we will be using is the Behavioral Risk Factor Surveillance System 
(BRFSS) data from the year [2013](https://www.kaggle.com/datasets/lplenka/brfss-data/data). Why are we using this dataset? I thought you’d never ask! The BRFSS is the actually nation’s premier system of health-related telephone surveys that collect state data about U.S. residents regarding their health-related risk behaviors, chronic health conditions, and use of preventive services. Established in 1984 with 15 states, BRFSS now collects data in all 50 states as well as the District of Columbia and three U.S. territories. 

BRFSS completes more than 400,000 adult interviews each year, making it the 
largest continuously conducted health survey system in the world. Impressive right? More information can be found [here.](https://www.cdc.gov/brfss/data_documentation/index.htm) 
The codebook for the 2013 data can be found [here.](https://www.cdc.gov/brfss/annual_data/2013/pdf/CODEBOOK13_LLCP.pdf)

In this data, there is a variable called **“General Health,”** which has posed the 
following questions to participants: *“Would you say that in general your health 
is…”* — 1) Excellent 2) Very good 3) Good 4) Fair or 5) Poor. This is a great example of Likert scale responses mentioned earlier. 


## Cleaning the Data to Create a Model Frame 

```{r}
#| label: dataset
#| echo: true
BRFSS_df1 <- read_csv("../data/05_brfss_subset.csv")
```

All things considered, the original is a somewhat large dataset, with 330 variables and 
having taken Microsoft Excel 2013 about three minutes to load the original csv. 
Patience is key! 

For this example, we are interested in the following variables:

**Independent Variables:**

*	Participant ID 
*	Sex 
*	Race 
*	Marital Status 
*	Highest Level of Schooling Completed 

**Dependent Variable:**

*	General Health  

The research question is straightforward: we want to know to what extent certain sociodemographic characteristics are associated with better self-reported health. 

```{r}
#| label: mutate
#| echo: true
BRFSS_df2 <- 
  BRFSS_df1 %>%
  as_tibble() %>% 
  drop_na(genhlth) %>% 
  mutate(
    Health = factor(genhlth, ordered = TRUE),
    Sex = factor(
      sex,
      levels = c("Male", "Female"),
      labels = c("Male", "Female")
    )
  ) %>%
  mutate(Race = as_factor(X_racegr3)) %>% 
  mutate(MaritalStatus = as_factor(marital)) %>% 
  mutate(
    Education = factor(
      educa,
      levels = c(
        "Never attended school or only kindergarten", 
        "Grades 1 through 8 (Elementary)", 
        "Grades 9 though 11 (Some high school)", 
        "Grade 12 or GED (High school graduate)", 
        "College 1 year to 3 years (Some college or technical school)",
        "College 4 years or more (College graduate)"
      ),
      labels = c(
        "Never attended school or only kindergarten", 
        "Grades 1 through 8 (Elementary)", 
        "Grades 9 through 11 (Some high school)", 
        "Grade 12 or GED (High school graduate)", 
        "College 1 year to 3 years (Some college or technical school)",
        "College 4 years or more (College graduate)"
      ),
      ordered = TRUE
    )
  ) %>% 
  mutate(Employment = as_factor(employ1)) %>% 
  mutate(
    Income = factor(
    income2,
    levels = c("Less than $10,000",
               "Less than $15,000",
               "Less than $20,000", 
               "Less than $25,000", 
               "Less than $35,000", 
               "Less than $50,000",
               "Less than $75,000",
               "$75,000 or more"),
    labels = c("Less than $10,000",
               "Less than $15,000",
               "Less than $20,000", 
               "Less than $25,000", 
               "Less than $35,000", 
               "Less than $50,000",
               "Less than $75,000",
               "$75,000 or more"),
    ordered = TRUE
    )
  ) %>%
  select(
    "Health","Sex", "MaritalStatus", "Education", "Income", "Employment", "Race"
  ) 
  
```

## Assumptions of the Method

The key assumptions of Ordinal logistic Regression which ensures the validity of 
the model are as follows: 

* The outcome variable is ordered.
* The predictor variables are either continuous, categorical, or ordinal.
* There is no multicollinearity among the predictors.
* Proportional odds.

In proportional odds logistic regression, the assumption is that the odds of 
being at or below any particular level of the ordinal dependent variable relative 
to being above that level is constant across levels of the independent variables. 
In other words, the relationship between the independent variables and the 
dependent variable is assumed to be proportional across all levels of the 
dependent variable. The model estimates coefficients for each independent 
variable, indicating how they influence the odds of being in a lower category 
versus a higher category of the [ordinal dependent variable](https://stats.oarc.ucla.edu/r/dae/ordinal-logistic-regression/). The interpretation 
of these coefficients is similar to that in binary logistic regression, but 
applies to each threshold between categories of the ordinal dependent variable.

One of the assumptions underlying ordinal logistic (and ordinal probit) 
regression is that the relationship between each pair of outcome groups is the 
[same](https://stats.oarc.ucla.edu/sas/dae/ordinal-logistic-regression/). In other words, ordinal logistic regression assumes that the coefficients 
that describe the relationship between, say, the lowest versus all higher 
categories of the response variable are the same as those that describe the 
relationship between the next lowest category and all higher categories, etc. 
This is called the proportional odds assumption or the parallel regression 
assumption. Because the relationship between all pairs of groups is the same, 
there is only one set of coefficients.

## Data Exploration

Lets take a look at the data we are working with. 

```{r}
#| label: dataskim
#| echo: true
# Skim Data
skim(BRFSS_df2)

BRFSS_df2 %>% 
  tbl_summary(by = Health) 
```

## Plotting Outcome Variable

```{r}
#| label: outcomeplot
#| echo: true
BRFSS_df2 %>% 
  count(Health) %>% 
  mutate(prop = n / sum(n)) %>% 
  rename(Health = Health) %>% 
  ggplot() + 
  aes(x = Health, y = prop) +
  labs(
    x = "Self Reported Health", 
    y = "Relative Frequencies (w Obs. Count)"
  ) +
  scale_y_continuous(labels = scales::percent) +
  geom_col() +
  geom_text(aes(label = n), vjust = 1.5, color = "white")

# Reorder due to outcome variable being ordered
BRFSS_df2 %>%
  ggplot(aes(x = fct_relevel(Health, c("Poor", "Fair", "Good", "Very good", "Excellent"))))+
  geom_bar() +
  labs(x = "Health")
```

```{r}
#| label: ggpairs
#| echo: true
# To plot this dataset, take a small sample (2000), due to it having a very large sample size
GGally::ggpairs(BRFSS_df2[sample(1:489790, size = 2000),])

```

```{r}
#| label: edulevel
#| echo: true
BRFSS_df3 <- 
  BRFSS_df2 %>% 
  select(-Income, -Employment) %>% 
  mutate(
    EducationLevel = case_when(
      Education == "Never attended school or only kindergarten" ~ 1,
      Education == "Grades 1 through 8 (Elementary)" ~ 1,
      Education == "Grades 9 through 11 (Some high school)" ~ 1,
      Education == "Grade 12 or GED (High school graduate)" ~ 2,
      Education == "College 1 year to 3 years (Some college or technical school)" ~ 3,
      Education == "College 4 years or more (College graduate)" ~ 4
    )
  ) %>% 
  mutate(
    EducationLevel = factor(
      EducationLevel, 
      levels = 1:4, 
      labels = c(
        "Less than HS", "HS/GED", "Some College", "College Grad"
      ), 
      ordered = TRUE
    )
  ) %>% 
  select(-Education)
```


At this point, we ought to consider the relationship between education, income, 
and employment, and whether or not it is best to choose one of these variables
as a predictor, as opposed to all three. Considering a majority of the literature 
highlights [education](https://www.sciencedaily.com/releases/2018/04/180416103428.htm) as one of the most important [predictors](https://archpublichealth.biomedcentral.com/articles/10.1186/s13690-020-00402-5) of health, we will
move forward with education only (out of the three aforementioned variables).

## Fitting the Model

Using the polr function, we can fit the Proportional Odds Logistic Regression 
model to the data. By default, the Hess option is turned off, but you can turn it on 
if you'd like to be able to calculate the odds ratios later.

```{r}
#| label: OLRFit
#| echo: true
olr_fit <- MASS::polr(Health ~ ., data = BRFSS_df3, Hess = TRUE)
```

The output is a bit messy, so it is wise to clean it up using the pander() 
function (from the pander:: and knitr:: packages).

```{r}
#| label: summaryOLRFit
#| echo: true
pander(summary(olr_fit))
```

## Interpreting the Regression Coefficients

**Sex:** Females were more likely than males to report higher self-reported health compared to males.
  
**Race:** Those who reported being white were significantly more likely to report better health when compared to black participants. Hispanics were more likely to report worse levels of health than those who were black. Those who reported being multiracial did not have a statistically significant difference from black participants. 
  
**Marital Status:** Compared to being divorced, those who reported being married and never having been married had greater levels of self-reported health. There was no statistical difference between those who reported being divorced and those who reported being separated. 
  
**Education:** The most notable statistically significant positive effect for greater self-reported health was present in the category where there was a change between those who fell under the category of having less than a high school education and having completed high school. For every increase in this particular educational category/level, the log odds of reporting a higher level of self-reported health increase by .15. 


## Generating P Values

Oftentimes, not having a p-value makes us itch with discomfort! One way to 
calculate a p-value here is by comparing the t-value against the standard normal 
distribution, like a z test. Note: this is only true with infinite degrees of 
freedom, but is reasonably approximated by large samples, becoming increasingly 
biased as sample size decreases. 

It is an easy and straightforward process: store the coefficient table, then 
calculate the p-values and combine back with the table.


```{r}
#| label: pvalues
#| echo: true

(ctable <- coef(summary(olr_fit)))

p <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
eBeta <- exp(ctable[, "Value"])

# Combine Table
ctable <- cbind(ctable, "expBeta" = eBeta)
(ctable <- cbind(ctable, "p value" = p))
```

## Conclusion

After running a statistical analysis consisting of an Ordinal Logistic Regression
for a dataset including an ordered outcome variable of self-reported health, we 
are able to see that being female, white, and being married all indicated a greater
likelihood of reporting higher levels of self-reported health. Additionally, certain 
changes in categories of education indicated greater odds of reporting greater
self-reported health. 
