---
title: "R Rayshader Overview"
subtitle: "PHC 6099 - Summer 2023"
author: "Kazi Tanvir Hasan"
toc: true
number-sections: true
format: html
embed-resources: false
---


## Overview

- R rayshader is an R package that allows users to generate high-quality 3D maps, visualizations, and animations.

- rayshader also allows the user to translate ggplot2 objects into beautiful 3D data visualizations.

To install rayshader, you can use the following code in R:

```{r}
#| message: false
# remotes::install_github(
#   "tylermorganwall/rayshader"
# )

# remotes::install_cran("rayrender") 
```


## Functions

- `rayshader` 0.35. 1 has 56 functions and 4 datasets

- seven functions related to mapping 

- also has functions to add water and generate overlays

- also included are functions to add additional effects and information to 3D visualizations

- functions for converting rasters to matrices

- functions to display and save your visualizations

- rayshader has a function to generate 3D plots using ggplot2 objects


## Example

First we load all the required libraries. These libraries are required for 
various functions and operations used in creating 3D maps with rayshader.

```{r}
#| message: false

library(rayshader)
library(rayrender) 
library(reshape2)
library(tidyverse)
```


## Example

Then, we download and load the data 

```{r}
#| eval: false

# Here, I load a map with the raster package.
loadzip <- tempfile() 

download.file("https://tylermw.com/data/dem_01.tif.zip", loadzip)

localtif <- raster::raster(
  unzip(loadzip, "dem_01.tif")
)

unlink(loadzip)

# write_rds(localtif, "../data/01_rayshader_eg_20240503.rds")
```

```{r}
#| echo: false
localtif <- read_rds("../data/01_rayshader_eg_20240503.rds")
```


In this code snippet, we create a temporary file (loadzip) to store the downloaded zip file from the specified URL. The `download.file()` function is used to download the file, and `unzip()` is used to extract the "dem_01.tif" file from the downloaded zip. Finally, we load the raster data into the localtif object.

## Create Map

To create a map first we need to convert this raster data file into a matrix 
using `raster_to_matrix()`

```{r}
#And convert it to a matrix:
elmat <- raster_to_matrix(localtif)
```

Then we use `sphere_shade()` and  `plot_map()` to create our base map

```{r}
elmat %>%
  sphere_shade(texture = "desert") %>%
  plot_map()
```

Here, elmat is a matrix created from the raster data using the `raster_to_matrix()` function. `sphere_shade()` applies shading to the elevation matrix, giving it a 3D effect. The texture parameter specifies the type of texture to be applied. In this case, it uses the "desert" texture. Finally, `plot_map()` is used to display the shaded map.


## Add Water Layer

We can add a water layer to the map using `detect_water()` and `add_water()`

```{r}
# detect_water and add_water adds a water layer to the map:
elmat %>%
  sphere_shade(texture = "desert") %>%
  add_water(detect_water(elmat), color = "desert") %>%
  plot_map()
```

In this code snippet, detect_water() function detects water areas in the elevation matrix. Then, `add_water()` adds a water layer to the map using the detected water areas. The color parameter specifies the color of the water. Finally, `plot_map()` is used to display the map with the water layer.


## Add Shadow Layer

We can also add shadow layer in the map.

```{r}
# And here we add an ambient occlusion shadow layer, which models lighting
#   from atmospheric scattering:

elmat %>%
  sphere_shade(texture = "desert") %>%
  add_water(detect_water(elmat), color = "desert") %>%
  add_shadow(ray_shade(elmat), 0.5) %>%
  add_shadow(ambient_shade(elmat), 0) %>%
  plot_map()
```

Here, `add_shadow()` is used to add a shadow layer to the map. `ray_shade()` calculates shadows based on the elevation matrix (elmat). The zscale parameter controls the strength of the shadows. `ambient_shade()` generates ambient lighting for the map. The second parameter of add_shadow() specifies the opacity of the shadows. Finally, `plot_map()` displays the map with shadows.


## Convert to 3D

We can convert this 2D map into 3D mapping using `plot_3d()` (by passing a texture map into the plot_3d function)

```{r}
elmat %>%
  sphere_shade(texture = "desert") %>%
  add_water(detect_water(elmat), color = "desert") %>%
  add_shadow(ray_shade(elmat, zscale = 3), 0.5) %>%
  add_shadow(ambient_shade(elmat), 0) %>%
  plot_3d(
    elmat, zscale = 10, fov = 0, theta = 135,
    zoom = 0.75, phi = 45, windowsize = c(1000, 800)
  )
Sys.sleep(0.2)
render_snapshot()
```

We can add a scale bar, as well as a compass using `render_scalebar()` and `render_compass()`

```{r}
render_camera(fov = 0, theta = 60, zoom = 0.75, phi = 45)
render_scalebar(
  limits = c(0, 5, 10),
  label_unit = "km",
  position = "W",
  y = 50,
  scale_length = c(0.33,1)
)
render_compass(position = "E")
render_snapshot(clear = TRUE)
```

Here, `render_camera()` sets the camera properties for the 3D map. `render_scalebar()` adds a scale bar to the map. The limits parameter specifies the limits of the scale bar, label_unit provides the label for the scale, position sets the position of the scale bar, y controls the vertical position, and scale_length determines the length of the scale bar. `render_compass()` adds a compass to the map, and `render_snapshot()` captures the final image of the map.


## 3D plotting with `rayshader` and `ggplot2`

Rayshader can also be used to make 3D plots out of ggplot2 objects using the `plot_gg()` function


```{r}
ggdiamonds = ggplot(diamonds) +
  stat_density_2d(
    aes(
      x = x, y = depth, fill = stat(nlevel)
    ), 
    geom = "polygon", n = 200, bins = 50,contour = TRUE
  ) +
  facet_wrap(clarity~.) +
  scale_fill_viridis_c(option = "A")

par(mfrow = c(1, 2))

plot_gg(ggdiamonds, width = 5, height = 5, raytrace = FALSE, preview = TRUE)

plot_gg(
  ggdiamonds, 
  width = 5, 
  height = 5, 
  multicore = TRUE, 
  scale = 250, 
  zoom = 0.7, 
  theta = 10, 
  phi = 30, 
  windowsize = c(800, 800)
)
Sys.sleep(0.2)
render_snapshot(clear = TRUE)
```


## Contour Plot 

Rayshader will automatically ignore lines and other elements that should not be mapped to 3D.

Here’s a contour plot of the volcano dataset.


```{r}
# Contours and other lines will automatically be ignored. Here is the volcano
#   dataset:

ggvolcano <- volcano %>% 
  melt() %>%
  ggplot() +
  geom_tile(aes(x = Var1, y = Var2, fill = value)) +
  geom_contour(aes(x = Var1, y = Var2, z = value), color = "black") +
  scale_x_continuous("X", expand = c(0, 0)) +
  scale_y_continuous("Y", expand = c(0, 0)) +
  scale_fill_gradientn("Z", colours = terrain.colors(10)) +
  coord_fixed()

par(mfrow = c(1, 2))
plot_gg(ggvolcano, width = 7, height = 4, raytrace = FALSE, preview = TRUE)
```
```{r}
plot_gg(
  ggvolcano,
  multicore = TRUE, 
  raytrace = TRUE, 
  width = 7, 
  height = 4, 
  scale = 300, 
  windowsize = c(1400, 866), 
  zoom = 0.6, 
  phi = 30, 
  theta = 30
)

Sys.sleep(0.2)

render_snapshot(clear = TRUE)
```



## `mtcars` Data Example

Rayshader also detects when the user passes the `color` aesthetic, and maps those values to 3D

```{r}
mtplot = ggplot(mtcars) + 
  geom_point(
    aes(x = mpg, y = disp, color = cyl)
  ) + 
  scale_color_continuous(limits = c(0, 8))

par(mfrow = c(1, 2))
plot_gg(mtplot, width = 3.5, raytrace = FALSE, preview = TRUE)

plot_gg(mtplot)
Sys.sleep(0.2)
render_snapshot(clear = TRUE)
```


## Reference

- <https://www.rayshader.com/>

- <https://www.youtube.com/watch?v=zgFXVhmKNbU>
