---
title: "Skimr Package"
author: "Shelly Sinclair and Alvonee Penn"
date: "`r Sys.Date()`"
format:
  html:
    self-contained: true
    toc: true
    number-sections: true
    code-fold: true
knitr:
  opts_chunk:      ########## set global options ############
    collapse: true # keep code from blocks together (if shown)
    echo: true    # don't show code
    message: false  # show messages
    warning: false  # show warnings
    error: false    # show error messages
    comment: ""    # don't show ## with printed output
    R.options:    
      digits: 3    # round to three digits
editor: source
bibliography: [Skimr_Package.bib]
---



# Introduction



`Skimr` is an R package designed to provide summary statistics about variables 
in data frames, tibbles, data tables and vectors. The function is modifiable 
where you can add additional variables, which are not a part of default summary 
function within R. Skimr allows us to quickly assess data quality by feature and 
type in a quick report. This is a critical step in Data Exploration, where 
Understanding our data helps us to generate a hypothesis and determine what data 
analysis are appropriate.

This presentation will cover the simplest and most effective ways to explore 
data in R.



##  Packages 



To begin we will upload the packages necessary for the lesson, this includes 
the following:

-   `readr()` to import our data file
-   `knitr()` that houses the `kable()` feature that allows us to construct and 
customize tables.
-   `tidyverse` houses the `dyplyr`package that assists with data manipulation 
and visualization.
-   The`skimr`package provides a compact summary of the variables in a dataset.

```{r}
#| label: install-Packages

# install.packages("skimr")
# install.packages("knitr")
# install.packages("tidyverse")

# load all the packages we will need to analyze the data and use the skim function
library(skimr)
library(knitr)
library(tidyverse)

```



## Census Data

For this assignment we will be using the Census_2010 dataset. There is no code 
book associated with the data, making it difficult to provide an accurate 
description of the variables. The information recorded shows the 
United States population estimates from the years 2010-2015, as well as relevant 
variables like net population change, number of births, number of deaths, 
international and domestic migration. Within the dataframe, there are 
3,193 observations and 100 variables.

The data can be imported into R from the following link: 
<https://fiudit-my.sharepoint.com/:x:/g/personal/ssinc013_fiu_edu/ESK1A13PstVGtf7HUwNNt68Bnh1YPfH8L-hnvMUxjBuCVw?e=CCwQU9>




```{r}
#| label: import-Data

# import the data
census_2010 <- read_csv("Data/census_2010.csv")

# what are the variables
colnames(census_2010) %>% 
  head(n = 10)

```



# The Summary() Function



In R, the most similar function is `summary()`. The `summary()` function in R 
can be used to quickly summarize the values in a data frame or vector[@Summary].

This syntax shows examples of the summary function using both our data set, and
a vector:

```{r}

#| label: Summary-syntax-with-data

# Example using summary function with data
summary(census_2010$CENSUS2010POP)

# Example using summary function with vector
# Define vector
x <- c(3, 4, 23, 5, 7, 8, 9, 12, 26, 15, 20, 21, NA)

# Summarize values in vector
summary(x)

```

The `summary()` function automatically calculates: The minimum value, The value 
of the 1st quartile (25th percentile), The median value, The value of the 3rd 
quartile (75th percentile) and The maximum value. Any missing values (NA) in the 
vector, the `summary()` function will automatically exclude them when 
calculating the summary statistics [@Summary].

Now, let's see how `skim()` compares.



# Skimr Package



The `skim()` function will generate a summary of the variables in your dataset, 
including their data type, number of non-missing values, minimum and maximum 
values, median, mean, standard deviation, and more [@Skimr].

The following syntax ensures that the data is compatible with `Skimr` 
functions.

```{r}
#| label: skimr-Package
#| code-fold: show

# is the summary data a skimr dataframe
skim(census_2010) %>% is_skim_df() # TRUE

```

We can explore the data as a tibble:

```{r}
#| label: create-Dataframe
#| code-fold: show

# use skim to get descriptive statistics of the data
skim(census_2010) %>% 
  head(n = 10)

```

Using `skimr` functions provides a cleaner and more detailed display of the 
results compared to the `summary()` function. In this example we are showing the 
first ten variables in our data set. The data summary tab shows the number of 
rows and columns, column type frequency and group variables. There is also 
additional descriptive information like missing values, unique characters. 

This will be relevant for data cleaning as well as understanding the distribution. 
Both are critical to determine which statistical analysis would be most 
appropriate to use for a project.



# Other Skimr Features



## Separate dataframes by type


The data frames produced by `skim()` are wide and sparse, filled with columns 
that are mostly NA. For that reason, it can be convenient to work with "by type" 
subsets of the original data frame. These smaller subsets have their NA columns 
removed.

Features:

* `partition()` - Creates a list of smaller data frames. Each entry in the list 
is a data type from the original dataframe
* `bind()` -  Takes the list and rebuilds the original dataframe.
* `yank()` - Extract a subtable from a dataframe with a particular type.

The following syntax is using `partition()` to separate the large census_df.

```{r}
#| label: separate-Data
#| code-fold: show

# split the character and numeric data
separate_df <- partition(skim(census_2010))
# check only the character data
separate_df$character

# create summary statistics for only numeric variables
numeric_separate_df <- separate_df[2]
# pull out the desired summary statistics in the nested list
head(numeric_separate_df$numeric["mean"]) %>% 
  kable(digits = 1) 

```

The following syntax is using `bind()` to combine the smaller character and 
numeric lists into the desired df.

```{r}
#| label: Bind-Data
#| code-fold: show

# combine the character and numeric data
head(bind(separate_df))

# confirm that the bound table is the same as the original skimmed table
identical(bind(separate_df), skim(census_2010)) 

```

The following syntax is using `yank()` to extract a specific table eg.character 
to examine.

```{r}
#| label: Yank-Data
#| code-fold: show

# Extract character data
yank(skim(census_2010), "character")

```


## Skimr with Dplyr


`Skimr` functions can be used in combination with `Dplyr` functions to examine 
specific variables within the census dataset.

The following example used `skim()` with `filter()` to display the variable 
`CENSUS2010POP`. The dataframe was further customized to display variable name 
and data type using `select()`.

```{r}
#| label: using-Dplyr
#| code-fold: show

# use dplyr functions on the statistics summary table
census_filter <- skim(census_2010) %>% filter(skim_variable == "CENSUS2010POP")
census_filter

census_select <- skim(census_2010) %>% select(skim_type, skim_variable)
head(census_select)

```

You can also customize the output of the `skim()` function by using various 
arguments. For example, you can use the numeric argument to specify which 
variables should be treated as numeric variables, or use the ranges argument to 
specify custom ranges for variables.

Using `skim()` in combination with `mutate()` we will compute a new variable to 
add to our skim dataframe.

```{r}
#| code-fold: show

# create a new variable calculate the change in birth rate from 2010 to 2011
census_2010 %>% 
  # new variable
  mutate(net_birth = BIRTHS2011 - BIRTHS2010) %>% 
  # move the variable to the beginning of the dataset
  relocate(net_birth, .after = CENSUS2010POP) %>% 
  # summary statistics table
  skim() %>% 
  # only the first fifteen variables
  head(n = 15) %>% 
  # change the formatting 
  kable(digit = 2)

```



## Adding Variables


* base	- An sfl that sets skimmers for all column types.
* append	- Whether the provided options should be in addition to the defaults 
already in skim. Default is TRUE.

As mentioned, `skim()` is designed to display default statistics, however you 
can use this function to change the summary statistics that it returns.

`skim_with()` is type closure: a function that returns adds a new variable to 
the table. This lets you have several skimming functions in a single R session, 
but it also means that you need to assign the return of `skim_with()` before 
you can use it.

You assign values within `skim_with()` by using the `sfl()` helper 
(skimr function list). It identifies which skimming functions you want to 
remove, by setting them to NULL. Assign an sfl to each column type that you 
wish to modify.

For example, we will add the following variables to the dataframe: 
median, min, max, IQR, length.


```{r}
#| label: add-Variables
#| code-fold: show

my_skim <- skim_with(numeric = sfl(median, min, max, IQR), character = sfl(length), 
            append = TRUE)

# add new variables into the summary table
census_2010 %>% 
  my_skim() %>% 
  head(n = 10)


```



# Conclusion



Overall, `Skimr` is a useful package for quickly summarizing the variables in a 
dataset and gaining insights into its structure and content.



# References 



